---
title: "Problem Set 1"
author: "Wen Sheng (ws2840)"
date: today
format:
  pdf:
    documentclass: article
    fontsize: 11pt
    geometry: margin=1in
    toc: false
    number-sections: false
execute:
  echo: true        # print code into the PDF
  warning: false
  message: false
  cache: true        # re-render fast; delete _cache/ to force a rerun
engine: knitr
editor:
  render-on-save: true
---

## Problem 1

### (i)

By Definition,

$$
\begin{align*}
E[Y|X=x] &= \sum_y y \times p(y|X=x) \\
&= \sum_y y \times \frac {P(X=x, Y=y)} {P(X=x)}
\end{align*}
$$

The joint probability distribution $P(X=x, Y=y)$ is:

$$
\begin{array}{c||ccccccccccc||c}
X \backslash Y
& 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12
& P(X=x) \\
\hline\hline
1
& \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
& \frac{1}{36} \\
2
& 0 & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
& \frac{3}{36} \\
3
& 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0
& \frac{5}{36} \\
4
& 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0
& \frac{7}{36} \\
5
& 0 & 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0
& \frac{9}{36} \\
6
& 0 & 0 & 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36}
& \frac{11}{36} \\
\hline\hline
P(Y=y)
& \frac{1}{36} & \frac{2}{36} & \frac{3}{36} & \frac{4}{36}
& \frac{5}{36} & \frac{6}{36} & \frac{5}{36} & \frac{4}{36}
& \frac{3}{36} & \frac{2}{36} & \frac{1}{36}
& 1
\end{array}
$$

So,
$$
\begin{align*}
E[Y|X=1] &= 2 * \frac {\frac{1}{36}}{\frac{1}{36}} = 2 \\ \\
E[Y|X=2] &= 3 * \frac {\frac{2}{36}}{\frac{3}{36}} + 4 * \frac {\frac{1}{36}}{\frac{3}{36}}= \frac {10} {3}\\ \\
E[Y|X=3] &= 4 * \frac {\frac{2}{36}}{\frac{5}{36}} + 5 * \frac {\frac{2}{36}}{\frac{5}{36}} + 6 * \frac {\frac{1}{36}}{\frac{5}{36}} = \frac {24} {5}\\ \\
E[Y|X=4] &= 5 * \frac {\frac{2}{36}}{\frac{7}{36}} + 6 * \frac {\frac{2}{36}}{\frac{7}{36}} + 7 * \frac {\frac{2}{36}}{\frac{7}{36}} + 8 * \frac {\frac{1}{36}}{\frac{7}{36}}= \frac {44} {7}\\ \\
E[Y|X=5] &= 6 * \frac {\frac{2}{36}}{\frac{9}{36}} + 7 * \frac {\frac{2}{36}}{\frac{9}{36}} + 8 * \frac {\frac{2}{36}}{\frac{9}{36}} + 9 * \frac {\frac{2}{36}}{\frac{9}{36}} + 10 * \frac {\frac{1}{36}}{\frac{9}{36}}= \frac {70} {9}\\ \\
E[Y|X=6] &= 7 * \frac {\frac{2}{36}}{\frac{11}{36}} + 8 * \frac {\frac{2}{36}}{\frac{11}{36}} + 9 * \frac {\frac{2}{36}}{\frac{11}{36}} + 10 * \frac {\frac{2}{36}}{\frac{11}{36}} + 11 * \frac {\frac{2}{36}}{\frac{11}{36}} + 12 * \frac {\frac{1}{36}}{\frac{11}{36}}= \frac {102} {11}
\end{align*}
$$

### (ii)

Simulate $n=1000$ draws of (X, Y). Plot the draws:

```{r, fig.dim = c(8, 8)}
set.seed(17)
n <- 1000
D1 <- sample(1:6, n, replace = TRUE)
D2 <- sample(1:6, n, replace = TRUE)

X <- pmax(D1, D2)
Y <- D1 + D2
ols <- lm(Y ~ X)

plot(X, Y, main = "Sum and Max of Two Dice Rolls")
abline(ols, col = "red", lwd = 2)
legend("topleft",
       sprintf("OLS: Y = %.2f + %.2fX", coef(ols)[1], coef(ols)[2]),
       col = "red", lwd = 2, bty = "n")
```

### (iii)

After simulation $n=1000$ observations and fitting the OLS regression,
$$
\hat{\beta}_0 = 0.34, \quad \hat{\beta}_1 = 1.50
$$

so

$$
\widehat{E[Y|X=x]} = \hat{\beta}_0 + \hat{\beta}_1 x = 0.34 + 1.50x
$$


### (iv)

### (v)


## Problem 2

## Problem 3

### (i)

The log-likelihood for $T$ i.i.d. observations is the following. We can prove the result.

### (ii)

### (iii)

The log-likelihood for $T$ i.i.d. observations is the following. We can prove the result.

### (iv)

## Problem 4

### (i)

The log-likelihood for $T$ i.i.d. observations is the following. We can prove the result.

### (ii)

## Problem 5

### (i)

The log-likelihood for $T$ i.i.d. observations is the following. We can prove the result.

### (ii)

## Problem 6

### (i)

The log-likelihood for $T$ i.i.d. observations is the following. We can prove the result.

### (ii)

### (iii)



```{r}
#| label: fig-airquality
#| fig-cap: "Temperature and ozone level."
#| warning: false

library(ggplot2)

ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")

```
