---
title: "Problem Set 1"
author: "Wen Sheng (ws2840)"
date: today
format:
  pdf:
    documentclass: article
    fontsize: 11pt
    geometry: margin=1in
    toc: false
    number-sections: false
execute:
  echo: true        # print code into the PDF
  warning: false
  message: false
  cache: true        # re-render fast; delete _cache/ to force a rerun
engine: knitr
editor:
  render-on-save: true
---

## Problem 1

### (i)

By Definition,

$$
\begin{align*}
E[Y|X=x] &= \sum_y y \times p(y|X=x) \\
&= \sum_y y \times \frac {P(X=x, Y=y)} {P(X=x)}
\end{align*}
$$

The joint probability distribution $P(X=x, Y=y)$ is:

$$
\begin{array}{c||ccccccccccc||c}
X \backslash Y
& 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12
& P(X=x) \\
\hline\hline
1
& \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
& \frac{1}{36} \\
2
& 0 & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0
& \frac{3}{36} \\
3
& 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0 & 0 & 0
& \frac{5}{36} \\
4
& 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0 & 0 & 0
& \frac{7}{36} \\
5
& 0 & 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36} & 0 & 0
& \frac{9}{36} \\
6
& 0 & 0 & 0 & 0 & 0 & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{2}{36} & \frac{1}{36}
& \frac{11}{36} \\
\hline\hline
P(Y=y)
& \frac{1}{36} & \frac{2}{36} & \frac{3}{36} & \frac{4}{36}
& \frac{5}{36} & \frac{6}{36} & \frac{5}{36} & \frac{4}{36}
& \frac{3}{36} & \frac{2}{36} & \frac{1}{36}
& 1
\end{array}
$$

So,
$$
\begin{align*}
E[Y|X=1] &= 2 * \frac {\frac{1}{36}}{\frac{1}{36}} = 2 \\ \\
E[Y|X=2] &= 3 * \frac {\frac{2}{36}}{\frac{3}{36}} + 4 * \frac {\frac{1}{36}}{\frac{3}{36}}= \frac {10} {3}\\ \\
E[Y|X=3] &= 4 * \frac {\frac{2}{36}}{\frac{5}{36}} + 5 * \frac {\frac{2}{36}}{\frac{5}{36}} + 6 * \frac {\frac{1}{36}}{\frac{5}{36}} = \frac {24} {5}\\ \\
E[Y|X=4] &= 5 * \frac {\frac{2}{36}}{\frac{7}{36}} + 6 * \frac {\frac{2}{36}}{\frac{7}{36}} + 7 * \frac {\frac{2}{36}}{\frac{7}{36}} + 8 * \frac {\frac{1}{36}}{\frac{7}{36}}= \frac {44} {7}\\ \\
E[Y|X=5] &= 6 * \frac {\frac{2}{36}}{\frac{9}{36}} + 7 * \frac {\frac{2}{36}}{\frac{9}{36}} + 8 * \frac {\frac{2}{36}}{\frac{9}{36}} + 9 * \frac {\frac{2}{36}}{\frac{9}{36}} + 10 * \frac {\frac{1}{36}}{\frac{9}{36}}= \frac {70} {9}\\ \\
E[Y|X=6] &= 7 * \frac {\frac{2}{36}}{\frac{11}{36}} + 8 * \frac {\frac{2}{36}}{\frac{11}{36}} + 9 * \frac {\frac{2}{36}}{\frac{11}{36}} + 10 * \frac {\frac{2}{36}}{\frac{11}{36}} + 11 * \frac {\frac{2}{36}}{\frac{11}{36}} + 12 * \frac {\frac{1}{36}}{\frac{11}{36}}= \frac {102} {11}
\end{align*}
$$

### (ii)

Simulate $n=1000$ draws of (X, Y). Plot the draws:

```{r, fig.dim = c(8, 8)}
set.seed(17)
n <- 1000
D1 <- sample(1:6, n, replace = TRUE)
D2 <- sample(1:6, n, replace = TRUE)

X <- pmax(D1, D2)
Y <- D1 + D2
ols <- lm(Y ~ X)

plot(X, Y, main = "Sum and Max of Two Dice Rolls")
abline(ols, col = "red", lwd = 2)
legend("topleft",
       sprintf("OLS: Y = %.2f + %.2fX", coef(ols)[1], coef(ols)[2]),
       col = "red", lwd = 2, bty = "n")
```

### (iii)

After simulation $n=1000$ observations and fitting the OLS regression,
$$
\hat{\beta}_0 = 0.34, \quad \hat{\beta}_1 = 1.50
$$

so

$$
\widehat{E[Y|X=x]} = \hat{\beta}_0 + \hat{\beta}_1 x = 0.34 + 1.50x
$$

\newpage

### (iv)

Using the estimated line, we get:

| x | True | OLS |
|---|---:|---:|
| 1 | 2.000 | 1.84 |
| 2 | 3.333 | 3.34 |
| 3 | 4.800 | 4.84 |
| 4 | 6.286 | 6.34 |
| 5 | 7.778 | 7.84 |
| 6 | 9.273 | 9.34 |

These estimates are **biased** becaused the true conditional expectation $E[Y|X=x]$ appears not linear on $X$, while the fitted OLS model assumed a linear relationship.

### (v)
Simulate and estimated B(x):

```{r}
M <- 2000
N <- 1000

x_vals <- 1:6
y_vals <- 2:12
join_prob <- matrix(0, nrow = 6, ncol = 11, dimnames = list(x_vals, y_vals))
join_prob["1", "2"] <- 1
join_prob["2", c("3", "4")] <- c(2, 1)
join_prob["3", c("4", "5", "6")] <- c(2, 2, 1)
join_prob["4", c("5", "6", "7", "8")] <- c(2, 2, 2, 1)
join_prob["5", c("6", "7", "8", "9", "10")] <- c(2, 2, 2, 2, 1)
join_prob["6", c("7", "8", "9", "10", "11", "12")] <- c(2, 2, 2, 2, 2, 1)
join_prob <- join_prob / 36

# calculate true E[Y|X=x]
P_X <- rowSums(join_prob)
true_EY_given_X <- as.vector(join_prob %*% y_vals) / P_X

# simulation: M=2000 times N=1000 draws
Yhat_mat <- matrix(NA, nrow = M, ncol = length(x_vals))
for (r in 1:M) {
  D1_r <- sample(1:6, N, replace = TRUE)
  D2_r <- sample(1:6, N, replace = TRUE)
  X_r <- pmax(D1_r, D2_r)
  Y_r <- D1_r + D2_r
  fit_r <- lm(Y_r ~ X_r)
  Yhat_mat[r, ] <- coef(fit_r)[1] + coef(fit_r)[2] * x_vals
}
EYhat <- colMeans(Yhat_mat)
Bias_x <- EYhat - true_EY_given_X
names(Bias_x) <- x_vals
bias_table <- data.frame(
x = x_vals,
true_EY = true_EY_given_X,
E_Yhat = EYhat,
Bias = Bias_x
)
bias_table

plot(x_vals, Bias_x, type = "b", pch = 19, col = "green",
xlab = "x", ylab = "Estimated bias B(x)",
main = "Estimated bias of the linear predictor Yhat(x)")
abline(h = 0, lty = 2, col = "gray50")
```

## Problem 2

Simplify the conditional density of $Y|X=x$:
$$
f(y|X=x) = \frac{f(x,y)}{f_X(x)}
$$ 

Let

$$
  z_x = \frac {x-\mu_x} {\sigma_x}, z_y = \frac {y-\mu_y} {\sigma_y}
$$

Then 
$$
\begin{align*}
f(x, y) &= \frac {1}{2 \pi \sigma_x \sigma_y \sqrt{1-\rho^2}}  \exp \left\{ -\frac {1} {1-\rho^2} \left[ z_x^2 - 2\rho z_x z_y + z_y^2 \right] \right\} \\
&= \frac {1}{2 \pi \sigma_x \sigma_y \sqrt{1-\rho^2}}   \exp \left\{ -\frac {1} {1-\rho^2} \left[ z_y^2 - 2\rho z_x z_y + \rho^2 z_x^2 - \rho^2 z_y^2 + z_x^2 \right] \right\} \\
&= \frac {1}{2 \pi \sigma_x \sigma_y \sqrt{1-\rho^2}}   \exp \left\{ -\frac {1} {1-\rho^2} (z_y - \rho z_x)^2 -\frac {1} {2} z_x^2 \right\} \\
&= \frac{1}{\sqrt{2\pi \sigma_y^2(1-\rho^2)}}   \exp \left\{ -\frac {1} {2} \left( \frac{z_y - \rho z_x}{\sqrt{1-\rho^2}} \right)^2 \right\} \frac{1}{\sqrt{2\pi \sigma_x^2}} \exp \left\{ -\frac {1} {2} z_x^2 \right\} \\
&= h(x, y) g(x)
\end{align*}
$$

Which $h(x,y)$ is also a valid normal density function:

$$
\begin{align*}
h(x, y) &= \frac{1}{\sqrt{2\pi \sigma_y^2(1-\rho^2)}}   \exp \left\{ -\frac {1} {2} \left( \frac{z_y - \rho z_x}{\sqrt{1-\rho^2}} \right)^2 \right\} \\
&= \frac{1}{\sqrt{2\pi \sigma_y^2(1-\rho^2)}} \exp \left\{ -\frac {1} {2} \left( \frac{\frac{y-\mu_y}{\sigma_y} - \rho \frac{x-\mu_x}{\sigma_x}}{\sqrt{1-\rho^2}} \right)^2 \right\} \\
&= \frac{1}{\sqrt{2\pi \sigma_y^2(1-\rho^2)}} \exp \left\{ -\frac {1} {2} \left( \frac{y-\mu_y - \rho \sigma_y\frac{x-\mu_x}{\sigma_x}}{\sigma_y \sqrt{1-\rho^2}} \right)^2 \right\} \\
\end{align*}
$$

The marginal distribution of X:

$$
\begin{align*}
f_X(x) &= \int_{-\infty}^{\infty} f(x, y) dy \\
&= \int_{-\infty}^{\infty} \frac{1}{\sqrt{2\pi \sigma_x^2}} \exp \left\{ -\frac {1} {2} z_x^2 \right\} h(x, y) dy \\
&= \frac{1}{\sqrt{2\pi \sigma_x^2}} \exp \left\{ -\frac {1} {2} z_x^2 \right\} \int_{-\infty}^{\infty} h(x, y) dy \\
&= \frac{1}{\sqrt{2\pi \sigma_x^2}} \exp \left\{ -\frac {1} {2} z_x^2 \right\}
\end{align*}
$$

Thefore, $f_X(x) = g(x)$

Now we have,

$$
\begin{align*}
f(y|X=x) &= \frac{f(x,y)}{f_X(x)} \\
&= \frac{h(x, y) g(x)}{g(x)} \\
&= h(x, y) \\
&= \frac{1}{\sqrt{2\pi \sigma_y^2(1-\rho^2)}} \exp \left\{ -\frac {1} {2} \left( \frac{y-\mu_y - \rho \sigma_y\frac{x-\mu_x}{\sigma_x}}{\sigma_y \sqrt{1-\rho^2}} \right)^2 \right\} \\
\end{align*}
$$

Hence,
$$
Y|X \sim N(\mu_y + \rho \frac{\sigma_y}{\sigma_x}(x-\mu_x), \sigma_y^2(1-\rho^2))
$$

## Problem 3

```{r}
#| echo: false
y <- c(199.0,205.0,196.0,200.0,218.0,220.0,215.0,223.0,237.0,234.0,235.0,230.0,250.0,248.0,253.0,246.0)
x <- c(rep(16,4),rep(24,4),rep(32,4),rep(40,4))   
```

### (i)
Fit OLS R Code:
```{r}
model <- lm(y ~ x)
model
```

### (ii)
We plot hardness against time for the 16 observations and overlay the fitted OLS regression line.

```{r, fig.dim = c(8, 5)}
plot(x, y, xlab = "X: time", ylab = "Y: hardness")
abline(model, col = "red", lty = 2)
legend("topleft",
       sprintf("OLS: Y = %.2f + %.2fX", coef(model)[1], coef(model)[2]),
       col = "red", lty = 2, bty = "n")
```

### (iii)
Compute MSE,
$$
 MSE = \frac{1}{n - 2} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 = 10.45893
$$

```{r}
y_hat <- fitted(model)
SSE <- sum((y - y_hat)^2)
MSE <- SSE / (length(y) - 2)
MSE
```

### (iv)

The sample correlation coefficient is:

$$
  corr = \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2 \sum_{i=1}^{n} (y_i - \bar{y})^2}} = 0.9865
$$
```{r}
cor(x, y)
```

The coefficeint of determination is:
$$
R^2 = 1 - \frac{SSE}{SST} = 1 - \frac{\sum_{i=1}^{n} (y_i - \hat{y}_i)^2}{\sum_{i=1}^{n} (y_i - \bar{y})^2} = 0.9731
$$
```{r}
SST <- sum((y - mean(y_hat))^2)
R2 <- 1 - SSE / SST
R2
```

## Problem 4

### (i)

Proof. The least-squares estimates $\hat \beta_0$ and $\hat \beta_1$ are obtained by minimizing the sum of squared residuals:
$$
Q(b_0, b_1) =\sum_{i=1}^{n} (y_i - b_0 - b_1 x_i)^2
$$

over $b_0, b_1 \in \mathbb{R}$. Because Q is differentiable on $\mathbb{R}^2$, the minimum occurs where the partial derivatives with respect to $b_0$ and $b_1$ are zero.
$$
\begin{align*}
0 &= \left. \frac{\partial Q}{\partial b_0}\right|_{b_0 = \hat{\beta}_0, b_1 = \hat{\beta}_1} \\
&= -2 \sum_{i=1}^{n} (y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i) \\
&= -2\sum_{i=1}^{n} e_i
\end{align*}
$$

Hence, the residuals sum to zero:

$$
\sum_{i=1}^{n} e_i = 0
$$

Similarly, differentiating with respect to the slope gives
$$
\frac{\partial Q}{\partial b_1} = -2 \sum_{i=1}^{n} x_i (y_i - b_0 - b_1 x_i) = 0
$$
Solving these equations gives the least-squares estimates:
$$
\begin{align*}
0 &= \left. \frac{\partial Q}{\partial b_1}\right|_{b_0 = \hat{\beta}_0, b_1 = \hat{\beta}_1} \\
&= -2 \sum_{i=1}^{n} x_i(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i) \\
&= -2\sum_{i=1}^{n} x_i e_i
\end{align*}
$$

Therefore,
$$
\sum_{i=1}^{n} x_i e_i = 0
$$
which proves part(i).

### (ii)

Proof. Use the fitted value for the $i$-th observation, we have
$$
\hat y_i = \hat{\beta}_0 + \hat{\beta}_1 x_i.
$$ 
and the two normal equations in Part (i). Then

$$
\begin{align*}
\sum_{i=1}^{n} \hat y_i e_i &= \sum_{i=1}^{n} (\hat{\beta}_0 + \hat{\beta}_1 x_i) e_i \\
&= \hat{\beta}_0 \sum_{i=1}^{n} e_i + \hat{\beta}_1 \sum_{i=1}^{n} x_i e_i \\
&= 0
\end{align*}
$$

Thus, the residual vector is orthogonal to the fitted-value vector, which proves part (ii).

## Problem 5

### (i)

Proof.

$$
\begin{align*}
\sum_{j=1}^n h_{ij}^2 &= \sum_{j=1}^n (\frac {1}{n} + \frac {(x_i - \bar{x})(x_j - \bar{x})}{S_{xx}})^2 \\
&= \sum_{j=1}^n (\frac {1}{n^2} + \frac{2(x_i - \bar{x})(x_j - \bar{x})}{n S_{xx}} + \frac{(x_i - \bar{x})^2 (x_j - \bar{x})^2}{S_{xx}^2})\\
&= \sum_{j=1}^n \frac{1}{n^2} + \frac{2(x_i - \bar{x})}{nS_{xx}}\sum_{j=1}^n (x_j - \bar{x})+ \frac{(x_i - \bar{x})^2}{S_{xx}^2} \sum_{j=1}^n(x_j - \bar{x})^2\\
&= \frac{1}{n} + 0 + \frac{(x_i - \bar{x})^2}{S_{xx}^2} S_{xx} \\
&= \frac{1}{n} + \frac{(x_i - \bar{x})^2}{S_{xx}} \\
&= h_{ii}
\end{align*}
$$

Hence, the sum of the squares of $i$-th row of the hat matrix: $\sum_{j=1}^n h_{ij}^2$ equals the diagonal element $h_{ii}$.

### (ii)

Proof.
$$
\begin{align*}
\sum_{j=1}^n h_{ij} x_j &= \sum_{j=1}^n (\frac {1}{n} + \frac {(x_i - \bar{x})(x_j - \bar{x})}{S_{xx}}) x_j \\
&= \frac{1}{n} \sum_{j=1}^n x_j + \frac{(x_i - \bar{x})}{S_{xx}} \sum_{j=1}^n (x_j - \bar{x}) x_j \\
&= \bar{x} +  \frac{(x_i - \bar{x})}{S_{xx}} S_{xx}\\
&= \bar{x} + (x_i - \bar{x}) \\
&= x_i
\end{align*}
$$

Hence, the dot product of the $i$-th row of the hat matrix with the vector of $x$ values: $\sum_{j=1}^n h_{ij} x_j$, equals the $i$-th $x$ value itself.

## Problem 6

### (i)

Under model (1),

$$
E[Y_i] = E[\beta x_i + \epsilon_i] = \beta x_i + E[\epsilon_i] = \beta x_i
$$

Then,
$$
\begin{align*}
E[\hat \beta] &= E\left[\frac{\sum_{i=1}^n x_i Y_i}{\sum_{i=1}^n x_i^2}\right] \\
&= \frac{\sum_{i=1}^n x_i E[Y_i]}{\sum_{i=1}^n x_i^2} \\
&= \frac{\sum_{i=1}^n x_i (\beta x_i)}{\sum_{i=1}^n x_i^2} \\
&= \frac{\beta \sum_{i=1}^n x_i^2}{\sum_{i=1}^n x_i^2} \\
&= \beta
\end{align*}
$$

Hence, $\hat \beta$ is unbiased for $\beta$.

### (ii)

### (iii)
